---
title: "Data wrangling with R - P2 Heatmap Generation"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Creates heatmap with virus levels. It assumes weekly data as the input.
It contains hardcoded scale for better visibility as the data have long tails in values

We continue our work with the data, now we will focus on generation of heatmap from the data_full.csv

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in waste water around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).


## Reading the data files after loading necessary libraries
 
Let's first load the necessary libraries and read the file data_full.csv file. As we know that the data_full.csv contains column names having "-", the option "check.names = F" is used to read it as it is. Otherwise the the hyphen is converted in to dot symbol while reading the file.

```{r}
library("tidyverse")
library("ComplexHeatmap")
library("circlize")
library("tidyr")
library("lubridate")
library("dplyr")


allData <- read.csv("../out/data_full.csv", check.names = F)
#calculated in scale-breakpoints.R
pos_start = 2.9
pos_end = 88
```

First we need to convert the "Date_collected" column in allData in to date format and extract the week and year information from that. The "as.Date" function from "lubridate" package can be used for this. 
We will then extract week from the date and populate in to a column "Week" in allData.The "sprintf" prints the week in 2 digit format. For eg. 1 is entered as 01. The "isoweek" is a function which converts date in to corresponding week. We have to also extract the Year from the date and populate it in to a column "Year" in the allData data frame   

```{r}
date =as.Date(allData$Date_collected, by="day")
allData$Week = sprintf("%02d", isoweek(date)) # Format week as 2 digit for sorting
allData$Month = sprintf("%02d", month(date)) # Format week as 2 digit for sorting
allData$Year = year(date)
head(allData)
```

As we can see two additional columns have been added as "Week" and "Year". Now let's use "group_by" function from "dplyr" library to group by columns: Site, Week and Year and add one more column "mean_size" containing mean of "Million_gene_copies_per_person_per_day". The data is grouped by Site, Week and Year and summarized based on mean of normalized data using pipe operator (%>%) in "dplyr" package.
Pipes let you take the output of one function and send it directly to the next, which is useful when you need to many things to the same data set. Pipes in R look like %>% and are made available via the "magrittr" package installed as part of "dplyr".

In the below code we use the pipe to send the allData set first through group_by() function, to get Site, Week and year grouped, and then through mutate() to add a new column "mean_size" containing mean of normalized data grouped by Site, Week and Year. The mutate() function adds new variables and preserves existing ones. So we are using mutate() instead of summarize() so that we can easily get which month each week falls (to be used in annotation) 
When the data frame is being passed to the group_by() and mutate() functions through a pipe, we don’t need to include it as an argument to these functions anymore.


```{r}
Meanaggregate<-allData %>% group_by(Site, Week,Year) %>% mutate(mean_size = mean(Million_gene_copies_per_person_per_day, na.rm = TRUE)) 
head(Meanaggregate)
```

Now let's use the "select" function from "dplyr" library to select Site, Month,Year, and mean_size from Meanaggregate and "unique" function to find the distinct rows.You can also do the selection using column numbers.


```{r}
Meanaggregate <- unique(select(Meanaggregate,Site,Month,Week,Year,mean_size)) 
head(Meanaggregate)
```

Let's sort the data frame Meanaggregate by ascending order of the months

```{r}
Meanaggregate<-Meanaggregate[order(Meanaggregate$Month),] #Sort by month
head(Meanaggregate)
```

We know that a week can be shared between two months. So it is necessary to set the month pointing to each week to starting month of that week for easier annotation.
The Meanaggregate data frame is grouped by Week and Year. The "mutate" function is used to change the Month to starting month of a week. 

After that we will select Site,Year,Month,Week and mean_size columns and take the unique rows to avoid same rows with duplicated months

```{r}

Meanaggregate<-Meanaggregate %>% group_by(Week,Year) %>% mutate(Month, Month=first(Month))
Meanaggregate<-unique(Meanaggregate) 
Meanaggregate<-as.data.frame(Meanaggregate) 
head(Meanaggregate)

```

We will now add "Year-Month-Week" column to the Meanaggregate data frame. The "paste()" function is used to paste Year , Month, and Week together by using seperator "-". Then the values are populated to the column "Year-Month-Week". We will then use the "select()" function to select columns Site, "Year-Moth-Week", and mean-size

```{r}
Meanaggregate$"Year-Month-Week" <- paste(Meanaggregate$Year,Meanaggregate$Month,Meanaggregate$Week,sep = "-")
Meanaggregate <- select(Meanaggregate, Site,"Year-Month-Week",mean_size)
```

Now we can convert the data to a wider format having column names from "Year-Month-Week" and values from the column mean_size

```{r}

Meanaggregate.wide <- pivot_wider(Meanaggregate, names_from = "Year-Month-Week", values_from = mean_size)

```

Let's convert the column "Site" to rownames for sorting and Sort the data ascending order based on the Column "Site"

```{r}
Meanaggregate.wide <- column_to_rownames(Meanaggregate.wide, var = "Site") 
Meanaggregate.wide <- Meanaggregate.wide[,order(colnames(Meanaggregate.wide))] # Sort ascending by column names
head(Meanaggregate.wide)
```
Now we will assign the column names of Meanaggregate.wide to Colnames. Then we will split the column names at "-" and put it in to three separate columns.  
The "within" function evaluates an R expression in an environment constructed based on a data frame AND modifies the original data. The "do.call" constructs and executes a function call from a name or a function and a list of arguments to be passed to it. In the below code the Colnames are splitted by "-" and binded together by calling "rbind" function with in dataframe environment of Colnames

```{r}
Colnames <- colnames(Meanaggregate.wide)
Colnames <- as.data.frame(Colnames)
Colnames1<-within(Colnames,Colnames<-data.frame(do.call('rbind', strsplit(as.character(Colnames), '-')))) 
```
We can convert the Second column(Colnames1$Colnames$X2) to month abbrevation and paste it to the first column(Colnames1$Colnames$X1) using "paste" function. Then we will take the unique dates for the annotation.

```{r}
Dates_Range <- paste(month.abb[as.integer(Colnames1$Colnames$X2)],as.integer(Colnames1$Colnames$X1),sep =  "-") -> Dates_Range # get the range of dates
unique(Dates_Range) -> Dates
Dates
```
Now Let's find out the annotation for the full Dates range

```{r}
c(1, head(cumsum(rle(Dates_Range)$lengths)+1, -1)) -> anno 
anno
```
Lets now find the date range for the selected Dates that is from Dec-2020 to Feb-2022

```{r}
DR = as.data.frame(Dates_Range)
selected_dates_range = DR[c(which(DR$Dates_Range=="Dec-2020")):(which(DR$Dates_Range=="Feb-2022")),] # extract the selected dates range
unique(selected_dates_range) -> selected_dates
selected_dates
```
We can find the annotation for selected Date range

```{r}
c(1, head(cumsum(rle(selected_dates_range)$lengths)+1, -1)) -> anno1 # find the annotation for selected date range
anno1
```

We need to extract the Data of interest from "Dec-2020" to "Feb-2022" 

```{r}
start_date_pattern = "2020-12-50" 
Meanaggregate.wide= as.data.frame(Meanaggregate.wide)
Meanaggregate.wide <- rownames_to_column(Meanaggregate.wide, var = "Site") 
start_date_cols = names(Meanaggregate.wide)
start_date_cols = start_date_cols[grepl(start_date_pattern, start_date_cols)]
data_selected = select(Meanaggregate.wide, Site | start_date_cols[1]:last_col())
```

Now we can select the sites which we are interested in

```{r}
selected_sites = c("Allanfearn","Helensburgh","Carbarns","Hamilton","Philipshill","Seafield","East Calder","Linlithgow","Shieldhall","Dalmuir","Paisley","Daldowie","Nigg","Peterhead","Kirkwall","Lerwick","Hatton","Forfar","Stirling","Falkirk","Alloa","Meadowhead","Stevenston","Linlithgow","Dunfermline","Levenmouth","Kirkcaldy","Troqueer","Lockerbie","Galashiels","Hawick","Stornoway")
data_selected = filter(data_selected, data_selected$Site %in% selected_sites)
```
We need to now do a custom ordering based on Health board

```{r}
ordered_sites = c("Allanfearn","Helensburgh","Carbarns","Hamilton","Philipshill","Seafield","East Calder","Linlithgow","Shieldhall","Dalmuir","Paisley","Daldowie","Nigg","Peterhead","Kirkwall","Lerwick","Hatton","Forfar","Stirling","Falkirk","Alloa","Meadowhead","Stevenston","Dunfermline","Levenmouth","Kirkcaldy","Troqueer","Lockerbie","Galashiels","Hawick","Stornoway")
data_selected = arrange(data_selected, factor(Site, levels = ordered_sites))
data_selected<-column_to_rownames(data_selected, var = "Site") 
Meanaggregate.wide<-column_to_rownames(Meanaggregate.wide, var = "Site")
```

We need to convert the data frames as matrix for passing in to the Heatmap function
```{r}
data_selected = as.matrix(data_selected)
data_full <- as.matrix(Meanaggregate.wide)
```

Specify the bottom annotation for heatmap
```{r}
ha = HeatmapAnnotation(foo=anno_mark(at = anno, labels = Dates,  labels_gp = gpar(fontsize=3, fontface = "bold"),which = "column", side = "bottom"))
ha1 = HeatmapAnnotation(foo=anno_mark(at = anno1, labels = selected_dates,  labels_gp = gpar(fontsize=3, fontface = "bold"),which = "column", side = "bottom"))
```

custom scale, positive start and end are used for rainbos range, non positive values are fixed to yellow

```{r}
col_fun1 = colorRamp2(c(0,pos_start,pos_end), c("yellow", "orange","blue")) 
```

```{r}
fName = "figure_prevalence_heatmap.jpeg"
jpeg(filename= fName, width=1000, height = 700, units= "px",res= 300) # Save as jpeg with width 1000 pixels, height = 700 pixels, and res = 300 dpi
hm = Heatmap(data_selected, col = col_fun1, row_names_gp = gpar(fontsize = 3, fontfamily = "Helvetica",fontface= "bold") , 
        column_names_rot = 90, 
        #column_title = "Covid virus prevelance at different sites along the timeline",
        #column_title_gp = gpar(fontsize = 5,fontface = "bold"), 
        na_col = "white", cluster_rows = FALSE, cluster_columns = FALSE , 
        bottom_annotation = ha1, show_column_names = FALSE, border = TRUE, border_gp = gpar(col = "grey"), 
        row_names_side = "left", 
        heatmap_legend_param = list(title="Virus levels\n[Mgc/(person*day)]", title_gp = gpar(fontsize =3, fontfamily = "Helvetica", fontface = "bold" ), at = c(0,pos_start,pos_end), 
                                    labels = c(0, "positive", pos_end), labels_gp= gpar(fontsize = 3, fontfamily = "Helvetica")))

draw(hm)
dev.off()
```
Let's now the heatmap for  full data

```{r}
fName = "figure_prevalence_heatmap_all.jpeg"
col_fun1 = colorRamp2(c(0,pos_start,pos_end), c("yellow", "orange","blue")) 
jpeg(filename= fName, width=1000, height = 2500, units= "px",res= 300) # Save as jpeg with width 1000 pixels, height = 700 pixels, and res = 300 dpi
hm = Heatmap(data_full, col = col_fun1, row_names_gp = gpar(fontsize = 3, fontfamily = "Helvetica",fontface= "bold") , 
        column_names_rot = 90, 
        na_col = "white", cluster_rows = FALSE, cluster_columns = FALSE , 
        bottom_annotation = ha, show_column_names = FALSE, border = TRUE, border_gp = gpar(col = "grey"), 
        row_names_side = "left", row_order = sort(rownames(data_full)),heatmap_legend_param = list(title="Virus levels\n[Mgc/(person*day)]", title_gp = gpar(fontsize =3, fontfamily = "Helvetica", fontface = "bold" ), at = c(0,pos_start,pos_end),labels = c(0, "positive", pos_end), labels_gp= gpar(fontsize = 3, fontfamily = "Helvetica")))
draw(hm)
dev.off()
```

