---
title: "Data wrangling with R - P2 Sampling sites file generation"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We continue our work with the data, now we will focus on Sampling sites file generation

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in waste water around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).

Firstly, read our renamed data.

## Reading the data files after loading necessary libraries

We need to read the curated_site_coordinate_data.csv and curated_population_data.csv files first

```{r}
library("lubridate")
library("tidyverse")

curated_data<- read.csv("../data/curated_data.csv", check.names = F)
sampling_sites<- read.csv("../out/sampling_sites.csv")
glimpse(curated_data)
glimpse(sampling_sites)

```
We can use command `glimpse` to see all the columns (variables) and first data in each column.
Now lets merge the two data sets by site

```{r}
merged<-merge(sampling_sites, curated_data, by = "Site")
head(merged)
```
We can see two Health_Board entries from both data sets. We need to keep only one of them as both are the same. Let's select the columns we need.


```{r}
merged_full<-select(merged,Health_Board.x,Site,Latitude_dd,Longitude_dd,Population,Population_Band,Date_collected,Date_collected,Date_analysed,SW_sample_number,N1_description,"N1_Reported_value-gc_per_L","N1_Repl_1-gc_per_L","N1_Repl_2-gc_per_L","N1_Repl_3-gc_per_L","Calculated_mean","Standard_Deviation","Flow-L_per_day","Ammonia-mg_per_L","pH_value","Modelled_flow-L_per_day","Million_gene_copies_per_person_per_day","Analysis_lab")
head(merged_full)
```
Now we need to rename Health_Board.x to Health_Board

```{r}
colnames(merged_full)[1] <- 'Health_Board'
```

Now the file is ready. We can create the out directory if it doesn't exist and save the file
```{r}
if (!file.exists('../out')) {
  dir.create(file.path('..', 'out'))  
}

fName = '../out/data_full.csv'
write.csv(merged_full, fName, row.names = FALSE)
```

