---
title: "Data wrangling with R - P2 Sampling sites file generation"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We continue our work with the data, now we will focus on Sampling sites file generation

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in waste water around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).

Firstly, read our renamed data.

## Reading the data files after loading necessary libraries

We need to read the curated_site_coordinate_data.csv and curated_population_data.csv files first

```{r}
library("lubridate")
library("sgo")
library("tidyverse")

site_coordinates<- read.csv("../data/curated_site_coordinate_data.csv")
site_population<- read.csv("../data/curated_population_data.csv")
glimpse(site_coordinates)
glimpse(site_population)

```
We can use command `glimpse` to see all the columns (variables) and first data in each column.
As we can see the Latitude_dd and Longitude_dd measurements are in different system (UK grid). Now we need to convert geographic locations into GPS DD unit
```{r}
locs <- sgo_points(site_coordinates, coords=c("Longitude_dd", "Latitude_dd"), epsg=27700)
Data_sgo_4326 <- sgo_bng_lonlat(locs, to=4326)
Data_4326 <- as.data.frame(Data_sgo_4326)
head(Data_4326)
```
The function sgo_points from "sgo" library converts UK grid system co-ordinates to GPS DD unit. We can see the Longitude_dd column is now named as x and Latitude_dd column as y. We need to rename those

```{r}
colnames(Data_4326) = c("Longitude_dd", "Latitude_dd", "Health_Board","Site")
head(Data_4326)
```
Now let's merge the coordinate information with population data and check how it looks

```{r}
merged <- merge( site_population, Data_4326,by = "Site")
head(merged)
```
As we know that there are Health_Board information in both site_population and Data_4326. So when these are merged we can see two columns Health_Board.x from site_population and Health_Board.y from Data_4326. We need to select one of these columns as both are the same.

```{r}
merged <- select(merged,Site,Health_Board.x,Latitude_dd,Longitude_dd,Population,Population_Band) 
head(merged)
```
Now let's rename the column Health_Board.x to Health.Board

```{r}
colnames(merged)[2] <- 'Health_Board'
head(merged)
```
Now the file is ready. We can create the out directory if it doesn't exist and save the file
```{r}
if (!file.exists('../out')) {
  dir.create(file.path('..', 'out'))  
}

fName = '../out/sampling_sites.csv'
write.csv(merged, fName, row.names = FALSE)
```

