---
title: "Data wrangling with R - P2 Data Curation"
author: "Sumy V Baby, Tomasz Zielinski"
date: "2022-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We continue our work with the data, now we will focus on actual data entries.

We are going to fix some of the typical problems with real data.

This and the following R-Markdown notebooks have been inspired by our real data curation work for the project about monitoring COVID virus levels in wastewater around Scotland ([see project page](https://biordm.github.io/COVID-Wastewater-Scotland/)).

Firstly, read our renamed data.

## Inspecting DATA

```{r}
library(dplyr)
renamed_data <- read.csv("../data/renamed_data.csv")
glimpse(renamed_data)

```
We used a different command `glimpse` to see all the columns (variables) and first data in each column.

Site stores information about where the samples were taken. As it is entered manually it is possible that it contains some misspelled locations.

## Chasing misspelled text

We can get all the unique values for the `Site` variable and sort them. That way we can inspect the values and try to find close by duplicates.



```{r}
sites = renamed_data$Site
sites = sort(unique(sites))
sites

```
There are **7**, sites with close but not identical names.  
One differs by having a space after the name.

Can you spot them?*

It would be better to identify them automatically.

We could count the measurements for each site and find those which have less then 5, those are probably misspelled as each site was monitored more than few times.

In order to do it .....

.... NOw we need a code that does that.


... then the code which takes a dictionary of wrong names and correct one and does the conversion
...
...


## The factors variables

Factors are variables that have only limitted set of options. 
In our case N1_description is such a variable, as it describes
the virust leves as Negative, Weak Poistive, Postives etc.

.... now the code to convert to factors and find the wrong factors.
((I change few descriptions to be misspelled (wrong letter size, mispelling or other key workd)))


Show how to easily find them, fix and convert to factors


## Dates

Convert to dates, represent in iso format, find the wrong date by sorting or min max (the 1899),
fix it.

## Number
Comment what to do with the numbers, for example use summary(Ammonia) to cehck it all look ok, 
negative value would probably be wrong.


Save the currated data to ../data/curated_data


------------
*The misspelled sites were: "Allanfearn ", "Invurie", "Langhlm", "Philiphill", "Sheildhall", "Stevenson", "hatton - Fintry West"